Extracting The Power of Dependent Types

### 2021-07-15 paper.tex
Do a (partial) pass over the paper

@@

It is often desirable to guarantee that a
certain class of errors does not occur in the given program.  These
error classes may include: dereferencing the null pointer, division
by zero, or out of bound indexing.  In most
programming languages, the guarantees that can be checked
statically are limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and/or libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

@@

To keep presentation as simple as possible, we have
chosen a simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
While the embedding may look simplistic, this simplicity
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

@@

\paragraph{Metaprogramming} There is a large body of
work on metaprogramming facilities in various programming
languages.  \citet{refl-masses} track the origins
of metaprogramming to Smith's work on reflection in Lisp~\cite{refl-lisp}.
Some prominent metaprogramming systems include
MetaOcaml~\cite{metaocaml}, MetaML~\cite{metaml},
reFlect~\cite{DBLP:journals/jfp/GrundyMO06},
Template Haskell~\cite{sheard2002template},
Racket~\cite{plt-tr1}, and various other Lisp/Scheme dialects.
%
However, these systems typically do not support dependent
types, so they are not well suited for our goal of statically
enforcing correctness of embedded programs.

\paragraph{Embedding}
Defining deep embeddings with static guarantees are a common application
of dependent types~\cite{10.5555/647849.737066,CHAPMAN200921,
10.1007/978-3-540-74464-1_7,10.1145/3236785,10.1145/1863495.1863497}.
%
These embeddings usually also define semantics of the embedded
language and therefore allow us to reason about the correctness
of program transformations and optimisations.
%
While the fact that this is possible is impressive in theory, the resulting
encodings are often difficult to use in practice. In this paper
we instead aim for a more lightweight approach.

\citet{deepshallow} propose to solve this problem with
a combination of deep and shallow embeddings.  Their idea is to define
a small deep embedding and leverage type classes in Haskell to define
the rest of the language on top of that.  It would be interesting to see
whether such an approach scales to dependently-typed embedded
languages.

\paragraph{Extraction}
The Coq proof assistant is equipped with extraction
capabilities~\cite{10.1007/978-3-540-69407-6_39,10.1007/3-540-39185-1_12},
which extracts functional code from Coq proofs (or programs).  The
default target language is Ocaml, but a few other options were added
recently.
%
Likewise, Agda itself has a mechanism for defining custom backends, of
which the GHC backend is the most prominent.
%
Other proof assistants provide similar extraction tools as well.
%
The main difference from our approach in this paper
is that these extractors are written as plugins to the proof
assistant, while we implement our extractors directly in the proof
assistant itself.
%
While it would be possible to implement extractors presented in this
paper as Coq plugins or Agda backends, conceptually they are more
heavyweight.  Our extractors and programs can (in principle)
communicate with each other. In addition, as they are just Agda programs, they can
be reflected themselves and their structure can be leveraged.

\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.

>>>>

\newcommand{\hrefu}[2]{\href{#1}{\nolinkurl{#2}}}

@@

It is often desirable to guarantee that a
certain class of errors does not occur in a given program.  These
error classes may include: dereferencing a null pointer, division
by zero, or out of bound indexing.  In most
programming languages, the properties that can be checked
statically are limited.  Nevertheless, moving to a different
language is not always an option.  For example, these languages may
come with great tooling, compilers and/or libraries.  In this paper, we
investigate how to provide strong static guarantees for programs in a
given language, while maintaining the ability to to use existing
compilers and tools.

@@

To keep presentation as easy as possible to follow, we have
chosen a simple target language to embed --- we use a subset
of PostScript~\citep{postscript1999language}, a stack-based
language used in the corresponding document format.
%
\figref{sqsum} shows an example of a program
written in our embedding and the PostScript code produced by
our extractor.
%
While the embedding may look simplistic, this simplicity
allows us to focus on the essence of the proposed approach and avoid unrelated
technical details.  We ensure that many verification patterns found in bigger
languages are covered, showing that the proposed approach scales.

@@

%\paragraph{Metaprogramming}
There is a large body of
work on metaprogramming facilities in various programming
languages.  \citet{refl-masses} track the origins
of metaprogramming to Smith's work on reflection in Lisp \cite{refl-lisp}.
Some prominent metaprogramming systems include
MetaOcaml \cite{metaocaml}, MetaML \cite{metaml},
reFlect \cite{DBLP:journals/jfp/GrundyMO06},
Template Haskell \cite{sheard2002template},
Racket \cite{plt-tr1}, and various other Lisp/Scheme dialects.
%
However, these systems typically do not support dependent
types, so they are not well suited for our goal of statically
enforcing correctness of embedded programs.

%\paragraph{Embedding}
Defining deep embeddings with static guarantees are a common application
of dependent types~\cite{10.5555/647849.737066,CHAPMAN200921,
10.1007/978-3-540-74464-1_7,10.1145/3236785,10.1145/1863495.1863497}.
%
These embeddings usually also define semantics of the embedded
language and therefore allow us to reason about the correctness
of program transformations and optimisations.
%
While the fact that this is possible is impressive in theory, the resulting
encodings are often difficult to use in practice. In this paper
we instead aim for a more lightweight approach.
%
\citet{deepshallow} propose to solve this problem with
a combination of deep and shallow embeddings.  Their idea is to define
a small deep embedding and leverage type classes in Haskell to define
the rest of the language on top of that.  However, so far this idea has not yet
been applied to dependently typed embedded languages.

%\paragraph{Extraction}
The Coq proof assistant is equipped with extraction
capabilities~\cite{10.1007/978-3-540-69407-6_39,10.1007/3-540-39185-1_12},
which extracts functional code from Coq proofs (or programs).  The
default target language is Ocaml, but a few other options were added
recently.
%
Likewise, Agda itself has a mechanism for defining custom backends, of
which the GHC backend is the most prominent.
%
Other proof assistants provide similar extraction tools as well.
%
The main difference from our approach in this paper
is that these extractors are written as plugins to the proof
assistant, while we implement our extractors directly in the proof
assistant itself.
%
While it would be possible to implement extractors presented in this
paper as Coq plugins or Agda backends, conceptually they are more
heavyweight.  Our extractors and programs can (in principle)
communicate with each other. In addition, as they are just Agda programs, they can
be reflected themselves and their structure can be leveraged.

%\paragraph{Dependently typed metaprogramming}
Several dependently-typed languages are equipped with metaprogramming
capabilities: Idris \cite{idris-refl}, Lean \cite{lean-refl},
Coq \cite{metacoq}, and Agda \cite{agda-refl}.  All of these
implement a similar API as described in this paper.  This is
reassuring, as it means our proposed approach is immediately portable
into many other contexts.
\citet{10.1145/3371071} introduce the Turnstile+ framework for
building dependently typed embedded DSLs and
shares the ideas advocated in this paper, suggesting that
our approach could work there as well.
\citet{10.1145/3371076} use MetaCoq to formally verify
the core type system of Coq. This combines nicely with
our approach, as we could use the verified core language
as a basis to verify our custom extractors.
\citet{10.1145/3372885.3373829} use MetaCoq to implement
a DSL combining deep and shallow approaches, in a way that
is quite similar to our own. While they are able to formally
reason about preservation of semantics (which we cannot do yet), it is unclear
whether their approach scales to dependently-typed embedded languages.
